---
title: "Introduction to Low Level R"
author: "Fabrice Rossi"
format: html
---

## Types and classes
### Types
```{r}
typeof(1.5)
typeof("toto")
typeof('also between single quotes')
typeof(TRUE)
typeof(1L)
typeof(1)
typeof(typeof)
typeof(list())
```

### Classes

```{r}
data(iris)
typeof(iris)
class(iris)
```

## Vectors
### Introduction
Many things in R are vectors (and most of the other things are lists).

A vector is a collection of values of the same type (possibly a single value).

```{r}
x <- 15.25
length(x) ## all vectors have a length
x[1]
```

```{r}
length(length(x))
typeof(length(x))
```

The `c` function can be seen as a creation function or, more accurately, as a
vector concatenation function. 

```{r}
y <- c(1, -2, 45, 1.5, -1e-5)
y
typeof(y)
y[2]
length(y)
```

```{r}
c(x, y)
```

### Character vectors

This is not R strength...

```{r}
z <- "Alice"
length(z)
```

Salvation comes from the `stringr` package. 

```{r}
library(stringr)
```

```{r}
str_length(z)
```

```{r}
complicated <- "😵 😡 😠 🤬 R是一门很好的编程语言 هي لغة برمجة رائعة"
complicated
length(complicated)
str_length(complicated)
```

```{r}
some_values <- c("😵 😡 😠 🤬", "R是一门很好的编程语言", "alice")
some_values
length(some_values)
str_length(some_values)
str_to_upper(some_values)
```

### Indexing

```{r}
x <- c(6, 5, 1, 2, 3, 4)
x[3]
```

```{r}
x[4] <- 15
x
```

Assignment can be understood has a value semantic assignment: `x` and `y` are
different vectors with the same initial content.

```{r}
y <- x
y
```

Therefore, modifying `x` has no effect on `y`.

```{r}
x[1] <- 18
x
y
```

Indexing is made with vectors of any length.

```{r}
x
x[c(2, 2, 1, 4, 3)]
```

```{r}
x[0] ## no possible meaning -> empty numeric vector
x[7] ## no value there -> not available
x[10]
```

```{r}
typeof(numeric(0))
length(numeric(0))
```

```{r}
typeof(NA)
length(NA)
```

```{r}
typeof(NA_integer_)
NA_integer_
```

```{r}
x
x[c(4, 1, 0, 15)]
```

```{r}
c(numeric(0), 5, numeric(0))
```

```{r}
x
x[c(-1, -4, -12)]
```


```{r}
x
x[c(1, 4)] <- c(-34, 42)
x
x[c(2, 10)] <- c(123, 78)
x
```

```{r}
y <- c("A", "B", "C")
y
y[c(TRUE, FALSE, TRUE)]
```
